---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# spbridge: Marginally interpretable spatial logistic regression with bridge processes

This is a R software repository corresponding to the manuscript

> "Marginally interpretable spatial logistic regression with bridge processes" by Changwoo Lee and David B. Dunson. arXiv preprint arXiv:2412.04744  https://arxiv.org/abs/2412.04744


## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r, eval = F}
# install.packages("devtools")
devtools::install_github("changwoo-lee/spbridge")
```


## Package manual

See spbridge_1.0.0.1.pdf for the package manual and example codes.
<!-- See R folder for the functions, which contains  -->

<!-- - bridge.R: univariate and multivariate bridge distribution and normal variance mixing distributions -->
<!-- - splogi_bridge.R: spatial logistic model with bridge process random effect, using Matern kernel -->
<!-- - splogi_bridge_lowrank.R: spatial logistic model with bridge process random effect, using low-rank Matern kernel -->
<!-- - splogi_gaussian.R: spatial logistic model with Gaussian process random effect, using Matern kernel -->
<!-- - splogi_gaussian_lowrank.R: spatial logistic model with Gaussian process random effect, using low-rank Matern kernel -->

<!-- badges: start -->
<!-- badges: end -->

## Motivation

 Consider the following logistic mixed model with a spatial random effect for spatially indexed binary data:
 
$$
\mathrm{logit}\{\mathrm{Pr}(Y_{ij} = 1 \mid x_{ij}, u(s_i))\} = x_{ij}^T\beta + u(s_i), \quad u(\cdot)\sim \text{ mean zero process}
$$

where $i=1,\dots,n$, $j=1,\dots,N_i$, $Y_{ij}$ is a binary response at location $i$ and individual $j$, $x_{ij}$ is a vector of covariates, $\beta$ is a vector of regression coefficients, and $u(s_i)$ is a spatial random effect at location $s_i$. When $N_i=1$ for all $i$, this reduces to point-referenced spatial binary data.

The random effect $u(\cdot)$, a mean zero stochastic process defined on the whole spatial domain, accounts for spatial dependence and allows for prediction at new location in a coherent probabilistic framework. The most popular choice of $u(\cdot)$ is the Gaussian process.

However, by adding the random effect, the log-odds interpretation of $\beta$ is changed from population-averaged to site-specific. In other words, $\beta$ now represents the change in the log-odds of $Y$ associated with a 1-unit increase of $x$, holding the site fixed. This is in contrast with population-averaged interpretation from the model $\mathrm{logit}\{\mathrm{Pr}(Y_{ij} = 1 \mid x_{ij})\} = x_{ij}^T\beta^M$ (as usual interpretation in logistic regression without random effects), where $\beta^M$ represents the change in the log-odds of $Y$ associated with a 1-unit increase of $x$ without conditioning on the site. When $u(\cdot)$ is a Gaussian process, the induced model for $\mathrm{logit}\{\mathrm{Pr}(Y_{ij} = 1 \mid x_{ij})\}$ is no longer a linear function of $x_{ij}$, and marginal interpretation of regression coefficient is lost.

We develop a marginally interpretable spatial logistic model for researchers who wish to **(1) have both population-averaged and site-specific log-odds ratio interpretation of regression parameters, (2) take account of spatial dependence, and (3) enable prediction of probability of $Y$ at a new location, all in a single modeling framework**.
Extending the work of [Wang and Louis (2003)](#references), we propose a novel stochastic process called bridge process for spatial random effects based on normal scale mixture that allows marginal interpretation of regression coefficients in spatial logistic regression. Under a Bayesian framework, the inference is carried out with highly efficient MCMC algorithm leveraging normal mixture representation and Polya-Gamma data augmentation [(Polson et al., 2013)](#references).


## Case study: Childhood malaria in the Gambia

We analyze childhood malaria prevalence data in the Gambia. 
It contains $\sum_{i=1}^n N_i = 2035$ children's malaria infection statuses from $n=65$ villages, along with covariates including age, bed net use, net treatment with insecticides, a satellite-derived measure of greenness, and the presence of a health center in the villages. The study of population-averaged associations between malaria prevalence and covariates is useful for nationwide decision-making; for example, on supplying bed nets. Prediction of malaria prevalence at new locations is also of substantial interest.

### 1. Data preparation

We first load data from geoR package, and standardize the data following [Gelman et al (2008)](#references).

```{r}
library(spbridge)
data(gambia)
N = length(gambia$pos) # 2035
y = gambia$pos # binary response, N = 2035 by 1 vector
# define id based on spatial coords unique values
id = as.numeric(factor(paste(gambia$x, gambia$y))) 
n = length(unique(id)) # 65
coords = unique(cbind(gambia$x, gambia$y)/1000) # n by 2 matrix, in km

# standardized covariates following Gelman et al (2008)
intercept = rep(1,N) # intercept
age = scale(gambia$age/365, scale = 2*sd(gambia$age/365)) # in years
netuse = gambia$netuse - mean(gambia$netuse)
treated = gambia$treated - mean(gambia$treated)
green = scale(gambia$green, scale = 2*sd(gambia$green))
green2 = scale(gambia$green^2, scale = 2*sd(gambia$green^2))
healthctr = gambia$phc - mean(gambia$phc)

X = cbind(intercept, age, netuse, treated, green, green2, healthctr)
colnames(X) = c("(Intercept)", "age", "netuse", "treated", "green", "green2", "healthctr")

centers = c(attr(age, "scaled:center"), mean(gambia$netuse), mean(gambia$treated), 
            attr(green, "scaled:center"), attr(green2, "scaled:center"), mean(gambia$phc))
scales = c(attr(age, "scaled:scale"), 1, 1, attr(green, "scaled:scale"), attr(green2, "scaled:scale"), 1)
  
```


### 2. Fit the model

Now we fit the spatial logistic model with bridge process random effect for inference of both $\beta^M$ and $\beta$, as well as Gaussian process random effects for comparison of site-specific estimate $\beta$. We closely followed the choice of prior suggested by [Gelman (2006)](#references) and [Gelman et al (2008)](#references), e.g. Cauchy prior for regression coefficients; see the manuscript for detailed settings. Running times are less than 2 mins for 11,000 iteration under Apple M1 CPU environment.

```{r, cache = T}
# spatial logistic model with bridge process random effect
fit_bridge = spbridge::splogi_bridge(y = y,
                   X = X,
                   id = id,
                   priors = list(beta_intercept_scale = 10,
                                 beta_scale = 2.5, beta_df = Inf, 
                                 rho_lb = 0.01, rho_ub = 100),
                   coords = coords,
                   smoothness = 0.5, nburn = 1000, nsave = 10000, nthin = 1)
fit_bridge$t_mcmc
# spatial logistic model with Gaussian process random effect
fit_gaussian = spbridge::splogi_gaussian(y = y,
                   X = X,
                   id = id,
                   priors = list(beta_intercept_scale = 10,
                                 beta_scale = 2.5, beta_df = Inf, 
                                 rho_lb = 0.01, rho_ub = 100),
                   coords = coords,
                   smoothness = 0.5, nburn = 1000, nsave = 10000, nthin = 1)
fit_gaussian$t_mcmc

```


### 3. Summarize the result

Now we summarize the MCMC results after back-transforming the data in the original scale. The conditional estimates are essentially identical. From the proposed model, we also obtain parameters with population-averaged interpretation as well that are comparable with results of [Cattelan and Varin (2018)](#references). The trace plot shows good convergence. 

```{r}
library(bayesplot)
scaleback <- function(x, centers, scales){
  c(x[1] - sum((x[-1] * centers) / scales),
    x[-1] / scales)
}

# summary from bridge process
beta_b_save = fit_bridge$post_save[,1:ncol(X)]
beta_b_save_orig = coda::mcmc(t(apply(beta_b_save, 1, scaleback, centers, scales)))
betam_b_save_orig = coda::mcmc(t(apply(fit_bridge$betam_save, 1, scaleback, centers, scales)))


summary(betam_b_save_orig)# population-averaged, bridge process random effect

summary(beta_b_save_orig)# site(subject)-specific, bridge process random effect 

# Compare with Gaussian process random effects model
beta_g_save = fit_gaussian$post_save[,1:ncol(X)]
beta_g_save_orig = coda::mcmc(t(apply(beta_g_save, 1, scaleback, centers, scales)))
summary(beta_g_save_orig)# site(subject)-specific, Gaussian process random effect
```

```{r}
# further mcmc summary examples
library(bayesplot)
mcmc_trace(fit_bridge$post_save)
mcmc_trace(fit_gaussian$post_save)
```



### References

- Cattelan, M. and Varin, C. (2018). Marginal logistic regression for spatially clustered binary
data. J. R. Stat. Soc. Ser. C. Appl. Stat., 67(4):939–959

- Gelman, A. (2006). Prior distributions for variance parameters in hierarchical models. Bayesian Anal., 1(3):515–534.

- Gelman, A., Jakulin, A., Pittau, M. G., and Su, Y.-S. (2008). A weakly informative default prior
distribution for logistic and other regression models. Ann. Appl. Stat., 2(4):1360–1383.

- Polson, N. G., Scott, J. G., and Windle, J. (2013). Bayesian inference for logistic models using
Pólya–Gamma latent variables. J. Am. Statist. Assoc., 108(504):1339–1349.

- Wang, Z. and Louis, T. A. (2003). Matching Conditional and Marginal Shapes in Binary Random
Intercept Models Using a Bridge Distribution Function. Biometrika, 90(4):765–775.
